---
title: "Reverse Engineering Post Story "
author: Wesley Brown, Kellina Wanteu, Audrey Decker, Rona Kobell
date: "10/31/2020"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

# In Baltimore, just a few years after police killed Freddie Gray, the city saw a surge in violence, with almost one homicide a day. During the same time, arrests dropped citywide by 14 percent. This story looks at the data to determine how Baltimore fits into other cities' crime patterns, and how the violence has loomed over the city and its survivors. Here is the story:

[As police struggle to solve homicides, Baltimore residents see an ‘open season for killing’](https://www.washingtonpost.com/investigations/as-police-struggle-to-solve-homicides-baltimore-residents-see-an-open-season-for-killing/2018/12/26/7ee561e4-fb24-11e8-8c9a-860ce2a8148f_story.html)

## Load libraries and settings

```{r}

# For general data analysis
library(tidyverse)

```

## Load and clean data 

DESCRIPTION OF WHAT THE DATA IS, WHERE YOU GOT IT, WHAT IT INCLUDES, INCLUDING A LIST OF WHAT EACH FIELD IS. A LINK TO IT. 

* uid | character | explanation of what this is
* reported_date | date | explanation of what it is

```{r}
# loading data
homicide_data <- read_csv("data/homicide-data.csv")

#cleaning data
homicide_data <- homicide_data %>%
  mutate(reported_date = as.character(reported_date)) %>%
  mutate(year = str_sub(reported_date, start = 1L, end = 4L))

# display the data below
homicide_data

```

## Analysis

### Finding 1

"As Baltimore has seen a stunning surge of violence, with nearly a killing each day for the past three years in a city of 600,000, homicide arrests have plummeted. City police made an arrest in 41 percent of homicides in 2014; last year (2017), the rate was just 27 percent, a 14 percentage point drop."

```{r}

# Filter for Baltimore only homicides 
baltimore_data <- homicide_data %>%
  filter(city == "Baltimore")

#### SSTILL NEED TO WRITE CODE, TO ANSWER THIS PART OF THE QUESTION
# with nearly a killing each day for the past three years in a city of 600,000 
  
# WRITE COMMENTS EXPLAINING STEP BY STEP WHAT'S HAPPENING

# create new object from Baltimore_data called Baltimore_data_year_2014_2017
# select for year, reported date, city, state, and deposition of the case
# filter for years 2014 to 2018
# group by how it was disposed of (disposition) and by year
# we counter the number of times it happened
# pivot: we made the table wider so that it can give us the different values for disposition (arrest, closed by arrest, open/no arrest, etc.)
# mutate by dividing the average value of closed_by_arrest and average value of closed_without_arrest to find the arrest percentage
#filter for years 2014 and 2017
#select only those two years

baltimore_data_year_2014_2017 <- baltimore_data %>%
  select(year, reported_date, city, state, disposition) %>%
  filter(year >= "2014") %>%
  filter(year <= "2018") %>%
  group_by(year, disposition) %>%
  count() %>%
  pivot_wider(names_from = disposition, values_from = n) %>%
  mutate(total_cases = `Closed by arrest` + `Closed without arrest` + `Open/No arrest`) %>%
  mutate(arrest_pct = `Closed by arrest` / total_cases * 100) %>%
  filter(year == 2014 | year == 2017)
  # SELECT ONLY TWO COLUMNS, YEAR, ARREST_PCT
 

baltimore_data_year_2014_2017


  
```

### Finding 2

"Of 50 of the nation’s largest cities, Baltimore is one of 34 where police now make homicide arrests less often than in 2014, according to a Washington Post analysis. In Chicago, the homicide arrest rate has dropped 21 percentage points, in Boston it has dropped 12 points and in St. Louis it is down 9."

**Discussion**: EXPLAIN WHY NUMBERS OR METHOD DON'T ADD UP

```{r}
#2) Of 50 of the nation’s largest cities, Baltimore is one of 34 where police now make homicide arrests less often than in 2014, according to a Washington Post analysis. 



#Which cities make less amount of arrest?

#comment out on steps
# take average amount of arrest from homicide data.
#select for year, city, state, and disposition
# filter for all years between 2014 and 2017
# group by city, year and disposition
# count
#mutate to new columns - total cases closed by arrest, closed without arrest, and open, no arrests for 2014 and 2017.
#pivot wider to get the value of arrest percent change
#mutate to create new column
# filter to percent change
#arrange as new columns

average_amount_of_arrest <- homicide_data %>%
  select(year, reported_date, city, state, disposition) %>%
  filter(year >= "2014") %>%
  filter(year <= "2017") %>%
  group_by(city, year, disposition) %>%
  count() %>%
  pivot_wider(names_from = disposition, values_from = n) %>%
  mutate_if(is.numeric , replace_na, replace = 0) %>%
  mutate(total_cases = `Closed by arrest` + `Closed without arrest` + `Open/No arrest`) %>%
  mutate(arrest_pct = `Closed by arrest` / total_cases * 100) %>%
  filter(year == 2014 | year == 2017) %>%
  select(city, year, arrest_pct) %>%
  pivot_wider(names_from = year, values_from = arrest_pct) %>%
  mutate(pct_change = `2014` - `2017`) %>%
  filter(pct_change > 0) %>%
  arrange(pct_change)
  ungroup() %>%
  count()

average_amount_of_arrest

# In Chicago, the homicide arrest rate has dropped 21 percentage points, in Boston it has dropped 12 points and in St. Louis it is down 9.

#comment out this finding
# take cities arrest rate and city homicide data
#select the year, the reported date, the city, the state, and the deposition
# filter for the years 2014 to 2017
#group by city, year, and deposition
#counted the average value
# pivot: made the table wider to show us values from disposition.
# mutate to replace N/A with 0
#mutate to create new columns for closed by arrest, closed w/o arrest and open no arrest
#filter again for years 2014 and 2017
# Selected the city, year, arrest percentage
# pivot to show years and balues from arrest percentage
# Mutuate the percentage change between 2014 and 2017
#filter again for cities Baltimore, Chicago and St. Louis
#create new frame called cities_arrest_rate
#display it
# Took those three cities and filtered to find the percentage change greater than 0
# Ungroup and count the values separately
# Find the three city rates

cities_arrest_rate <- homicide_data %>%
  select(year, reported_date, city, state, disposition) %>%
  filter(year >= "2014") %>%
  filter(year <= "2017") %>%
  group_by(city, year, disposition) %>%
  count() %>%
  pivot_wider(names_from = disposition, values_from = n) %>%
  mutate_if(is.numeric , replace_na, replace = 0) %>%
  mutate(total_cases = `Closed by arrest` + `Closed without arrest` + `Open/No arrest`) %>%
  mutate(arrest_pct = `Closed by arrest` / total_cases * 100) %>%
  filter(year == 2014 | year == 2017) %>%
  select(city, year, arrest_pct) %>%
  pivot_wider(names_from = year, values_from = arrest_pct) %>%
  mutate(pct_change = `2014` - `2017`) %>%
  #filter(city == "Chicago" | city =="Baltimore" | city=="St. Louis") 
  filter(city %in% c("Chicago","Baltimore", "St. Louis"))
cities_arrest_rate

cities_arrest_less <- city_arrest_rate_change %>%
  filter(pct_change > 0) %>%
  ungroup() %>%
  count()

# Three city arrest rate

three_city_rate <- city_arrest_rate_change %>%
   filter(city %in% c("Chicago","Baltimore", "St. Louis"))

```



### Finding 3

Baltimore is also one of 30 cities that have seen an increase in homicides in recent years, with the greatest raw number increase in killings of any city other than Chicago, which has four times the population.

```{r}

# Baltimore is also one of 30 cities that have seen an increase in homicides in recent years.

city_homicide_change <- homicide_data %>%
  group_by(city, year) %>%
  summarise(number_homicides = n()) %>%
  filter(year == "2011" | year == "2017") %>%
  pivot_wider(names_from = year, values_from = number_homicides) %>%
  mutate(homicide_change = `2017` - `2011`) %>%
  filter(homicide_change > 0) %>%
  arrange(desc(homicide_change))

#display below
city_homicide_change

# Baltimore has had the greatest raw number increase in killings of any city but Chicago. 

#create new column of raw number of homicides
#compare from 2011 to 2017 to show an increase
#produce Chicago raw killings
#produce Baltimore raw killings

city_number_homicides <- homicide_data %>%
  group_by(city, year) %>%
  summarise(number_homicides = n()) %>%
  filter(year == "2011" | year == "2017") %>%
  pivot_wider(names_from = year, values_from = number_homicides) %>%
  #filter(city == "Chicago" | city =="Baltimore")
  filter(city %in% c("Chicago","Baltimore")) %>%
  arrange(desc(homicide_change))

#display below
city_number_homicides


#Chicago population is 4 times that of Baltimore
#WHERE DO WE GET POPULATION INFO
#comment out this finding
  
# We created a new data and named it city homicide change
  # then we grouped the desired data by city and year 
  # we filtered the year 2011 until 2017, eveything in between
  # we increased the number of columns and decreased the number of rows by using the pivot wider function. We took the names from those years listed above and the values from the number of homicides
  # then we created a new variable that would give us the percent change in homide from 2011 to 2017
  # We filtered the ones that had a change greater than 0
  # We arranged our new data from lowest to highest

# Baltimore has had the greatest raw number increase in killings of any city but Chicago. 
  
  #We filtered the data by choosing only the city of Chicago
  #Ten we filtered the average between 2007 and 2018
  # We did the same thing for Baltimore 
  # We compared the change between the two city to see which one had a greater increase in killings
```

```{r}
#4) While there is evidence for and against a nationwide Ferguson effect — the theory that crime increased after 2014 as police faced more scrutiny following the shooting of Michael Brown in Ferguson, Mo. — in Baltimore there is an indisputable Freddie Gray effect. As violence in the city has risen since 2015, the likelihood of a killer being arrested has dropped precipitously.


# Group by year and disposition
# Summarized the total number of values within those two categories
# Pivot wider to see total number from different disposition 
# Mutate to find total cases by adding values from close by arrest and closed without arrest and open/no arrest
# Mutate to find arrest percentage by divinding closed arrest by the total case we found in the previous line
#Filter from year 2014 to 2017
# Select year, total cases and arrest percentage change
baltimore_data_year_2014_2017 <- baltimore_data %>%
  group_by(year, disposition)  %>%
  summarise(total_number=n()) %>%
pivot_wider(names_from = disposition, values_from = total_number) %>%
  mutate(total_cases = `Closed by arrest` + `Closed without arrest` + `Open/No arrest`) %>%
  mutate(arrest_pct = `Closed by arrest` / total_cases * 100) %>%
  filter(year == 2014 | year == 2017) %>%
select(year, total_cases, arrest_pct)

# Comments from Sean
# year | arrests | killings | pct_arrested 
# 2014
# 2017
#change the number and show how it's different. 2014 and 2015
#there were other factors in the change in homicides
#is it choosing the time range?
#depending on when you start your query, your results can change so think about why you are making that decision.
```